\texttt{MSnbase}, efficient R-based processing and visualisation of raw mass spectrometry data

### 2020-04-28 msnbase2.Rnw
Smaller fixes and reformulations

\author{Laurent Gatto}
\email{laurent.gatto@uclouvain.be}
\affiliation[UCLouvain]{Computational Biology Unit, de Duve Institute, Universit\'e catholique de Louvain, Brussels, Belgium}
\author{Sebastian Gibb}
\affiliation[University of Greifswald]{Department of Anaesthesiology and Intensive Care of the University Medicine Greifswald, Germany}
\author{Johannes Rainer}
\affiliation[EURAC]{Institute for Biomedicine, Eurac Research, Affiliated Institute of the University of L\"ubeck, Bolzano, Italy}

@@

Mass spectrometry is a powerful technology to assays chemical and
biological samples. It is used routinely with well characterised
protocol, as well a development platform, to improve on existing
methods and devise new ones. The complexity and diversity of mass
spectrometry yields data that is itself complex and often times of
considerable size, that requires non trivial processing before
producing interpretable results. This is particularly relevant, and
can constitute a significant challenge for method developers: in
addition to the development of sample processing and mass spectrometry
methods, there is a need to process, analyse, interpret and assess
these new data to demonstrate the improvement in their technical and
analytical workflows.

There exists a very diverse catalogue of software tools to explore,
process and interpret mass spectrometry data. These range from low
level software libraries that are aimed at programmers to develop new
applications, to user-oriented applications. Example of software
libraries include Java-based jmzML~\cite{Cote:2010} or C/C++-based
ProteoWizard~\cite{Chambers:2012}, that are used as application
programming interfaces. ProteomeDiscoverer (Thermo Scientific),
MaxQuant~\cite{Cox:2008} and PeptideShaker~\cite{Vaudel:2015} are
among the most widely user-centric application. They generally provide
a limited and fixed set of functionality featuring often advanced
graphical user interfaces.

In this software note, we present version 2 of the \texttt{MSnbase}
\cite{Gatto:2012} software package, available from the
Bioconductor~\cite{Huber:2015} project. \texttt{MSnbase}, like other
Python-based software packages such as {pyOpenMS}~\cite{Rost:2014},
spectrum\_utils~\cite{Bittremieux:2020} or
Pyteomics~\cite{Goloborodko:2013}, offers a platform that lies between
low level libraries and end-use software. \texttt{MSnbase} provides a
flexible R~\cite{R} command line environment for metabolomics and
proteomics mass spectrometry-based application. It allows a detailed
step-by-step processing, analysis and exploration of the data and
development of novel computational mass spectrometry methods.

@@

In the following sections, we will describe demonstrate how
\texttt{MSnbase} effichiently manages large raw mass spectrometry data
and how it is used for data processing and visualisation. We will also
illustrate how it makes use of the forward-pipe operator
(\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
operator has proved useful to develop non-trivial analyses by
combining individual functions into easily readable pipelines.

@@

The main feature in version 2 of the \texttt{MSnbase} package was the
addition of different backends for raw data storage, namely
\textit{in-memory} and \textit{on-disk}. The following code chunk
demonstrates how to create two \texttt{MSnExp} objects, tailored to
manage mass spectrometry experiments, storing data in-memory or
on-disk.

@@

The former is the legacy storage mode, implemented in the first
version of the package, that loads all the raw data and the metadata
in memory. This solution doesn't scale for modern large dataset, and
was complemented by the on-disk backend, that only loads metadata into
memory and accesses the spectra in the original files when
needed. There are two direct benefits using the on-disk backend,
namely faster reading and reduced memory footpring. Figure
\ref{fig:bench} shows 5-fold faster reading times (a) and over a
10-fold reduction in memory usage (b).

The on-disk backend also offers efficient data manipulation by way of
\textit{lazy processing}. Operations on the raw data are stored in a
processing queue and only effectively applied when raw data is
accessed on disk. As an example, the following short analysis
pipeline, that can equally be applied to on in-memory or on-disk data
retains MS2 spectra acquired between 1000 and 3000 seconds, extract
the m/z range corresponding to the TMT 6-plex range and focuses on the
MS2 spectra with a precursor intensity greater than $11 \times 10^6$
(the median precursor intensity).

@@

As shown on Figure~\ref{fig:bench} (c), this lazy mechanism si
significantly faster than its application on in-memory data. The
advantageous reading and execution times and memory footprint of the
on-disk backend are possible by avoiding unnecessary access to the raw
data. Once access to the spectra m/z and intensity values become
mandatory (for example for plotting), then the in-memory backend
becomes more efficient, as illustrated on Figure~\ref{fig:bench}
(d). This gain is maximal when the whole dataset is the be accessed
(i.e. all spectra are already in memory) and negligeable when large
fractions of the data need to be subset.

This new on-disk infrastructure enables large scale data analyses
using \texttt{MSnbase} (metabolomics example, see Johannes).

>>>>

\author{Laurent Gatto}
\email{laurent.gatto@uclouvain.be}
\affiliation[UCLouvain]{Computational Biology Unit, de Duve Institute, Universit\'e catholique de Louvain, Brussels, Belgium}
\author{Sebastian Gibb}
\affiliation[University of Greifswald]{Department of Anaesthesiology and Intensive Care of the University Medicine Greifswald, Germany}
\author{Johannes Rainer}
\affiliation[Eurac Research]{Institute for Biomedicine, Eurac Research, Affiliated Institute of the University of L\"ubeck, Bolzano, Italy}

@@

Mass spectrometry is a powerful technology to assay chemical and
biological samples. It is used routinely with well characterised
protocol, as well a development platform, to improve on existing
methods and devise new ones. The complexity and diversity of mass
spectrometry yields data that is itself complex and often times of
considerable size, that requires non trivial processing before
producing interpretable results. This is particularly relevant, and
can constitute a significant challenge for method developers: in
addition to the development of sample processing and mass spectrometry
methods, there is a need to process, analyse, interpret and assess
these new data to demonstrate the improvement in their technical and
analytical workflows.

There exists a very diverse catalogue of software tools to explore, process and
interpret mass spectrometry data. These range from low level software libraries
that are aimed at programmers to enable development of new applications, to more
user-oriented applications with graphical user interfaces. Example of software
libraries include Java-based jmzML~\cite{Cote:2010} or C/C++-based
ProteoWizard~\cite{Chambers:2012}, that are used as application programming
interfaces. ProteomeDiscoverer (Thermo Scientific), MaxQuant~\cite{Cox:2008} and
PeptideShaker~\cite{Vaudel:2015} are among the most widely used user-centric
applications. They generally provide a limited and fixed set of functionality
featuring often advanced graphical user interfaces.

In this software note, we present version 2 of the \texttt{MSnbase}
\cite{Gatto:2012} software package, available from the
Bioconductor~\cite{Huber:2015} project. \texttt{MSnbase}, like other
software packages such as Python-based {pyOpenMS}~\cite{Rost:2014},
spectrum\_utils~\cite{Bittremieux:2020} or
Pyteomics~\cite{Goloborodko:2013}, offers a platform that lies between
low level libraries and end-use software. \texttt{MSnbase} provides a
flexible R~\cite{R} command line environment for metabolomics and
proteomics mass spectrometry-based applications. It allows a detailed
step-by-step processing, analysis and exploration of the data and
development of novel computational mass spectrometry methods~\cite{Stanstrup:2019}.

@@

In the following sections, we describe \texttt{MSnbase}'s ability to efficiently
manage very large mass spectrometry data and experiments and how it can be used
for data processing and visualisation. We will also illustrate how it makes use
of the forward-pipe operator (\texttt{\%>\%}) defined in the \texttt{magrittr}
package. This operator has proved useful to develop non-trivial analyses by
combining individual functions into easily readable pipelines.

@@

The main feature in version 2 of the \texttt{MSnbase} package was the
addition of different backends for raw data storage, namely
\textit{in-memory} and \textit{on-disk}. The following code chunk
demonstrates how to create two \texttt{MSnExp} objects, tailored to
manage mass spectrometry experiments, storing data either in-memory or
on-disk.

@@

The former is the legacy storage mode, implemented in the first
version of the package, that loads all the raw data and the metadata
into memory. This solution doesn't scale for modern large dataset, and
was complemented by the on-disk backend, that only loads metadata into
memory and accesses the spectra data (i.e. m/z and intensity values) in the original files when
needed. There are two direct benefits using the on-disk backend,
namely faster reading and reduced memory footprint. Figure
\ref{fig:bench} shows 5-fold faster reading times (a) and over a
10-fold reduction in memory usage (b).

The on-disk backend also offers efficient data manipulation by way of
\textit{lazy processing}. Operations on the raw data are stored in a
processing queue and only effectively applied when the user accesses m/z or
intensity values. As an example, the following short analysis
pipeline, that can equally be applied to in-memory or on-disk data,
retains MS2 spectra acquired between 1000 and 3000 seconds, extracts
the m/z range corresponding to the TMT 6-plex range and focuses on the
MS2 spectra with a precursor intensity greater than $11 \times 10^6$
(the median precursor intensity).

@@

As shown on Figure~\ref{fig:bench} (c), this lazy mechanism is
significantly faster than its application on in-memory data. The
advantageous reading and execution times and memory footprint of the
on-disk backend are possible by retrieving only spectra data from the selected
subset hence avoiding access to the full raw
data. Once access to the spectra m/z and intensity values become
mandatory (for example for plotting), then the in-memory backend
becomes more efficient, as illustrated on Figure~\ref{fig:bench}
(d). This gain is maximal when the whole dataset is accessed
(i.e. all spectra are already in memory) and negligible when only subsets of the
data are requested.

This new on-disk infrastructure enables analyses also of very large scale
experiments using \texttt{MSnbase}). An example analysis of an intermediate
scale metabolomics experiment (~2000 raw data files) using the
\texttt{xcms}~\cite{Smith:2006} R package, that completely builds on the on-disk infrastructure
of \texttt{MSnbase}, is described in [TODO: add citation once published.]
